\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-01 paper.qmd
small tweaks to notation and start on methodology

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$\ }
  \newcommand\kD{$k\text{-}D$\ }
---

@@

At first glance, thinking of of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The \pD observations are the realization of the phenomenon, and the \kD NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD layout to be variance that the model explains, and the (relative) difference with their distances in \pD is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in \pD that can be compared with the observed value.

### Notation

@tbl-notation summarises the notation used to explain the new methodology.

@@

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("5cm", "12cm"))
```

Once we have notation, we need to have some math that gives precision to the paragraph above. (XXX Michael??)

>>>>

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$\ }
  \newcommand\kD{$k\text{-}D$\ }
  \newcommand\dD{$d\text{-}D$\ }
---

@@

At first glance, thinking of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The \pD observations are the realization of the phenomenon, and the \kD NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD layout to be variance that the model explains, and the (relative) difference with their distances in \pD is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in \pD that can be compared with the observed value.

<!--### Notation -->

@tbl-notation summarises the notation used to explain the new methodology. $X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$ represents the observed data matrix, that are $n$ \pD vectors sampled from some population.

<!--
$X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$

$Y_{n \times d} = \begin{bmatrix} \textbf{y} _{1} & \textbf{y}_ {2} & \cdots & \textbf{y}_{n} \\  \end{bmatrix}^\top$

$C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$

$C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$ $p$-D mappings of 2D hexagonal bin centroids of the $k^{th}$ hexagon

-->

@@

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("3cm", "12cm"))
```

We need to have some math that gives precision to the paragraphs in the previous section, defining mapping from \pD to \kD to get the \kD layout. Might be written below, too (XXX Michael??)

To illustrate the method, we use simulated data. The S-curve is generated as follows XXX Jayani

### Generating the \kD layout

The usual procedures are followed to produce a \kD layout, using the preferred method, t-SNE, UMAP, ... The data may need to be pre-processed by standardising or transforming to principal components.

### 2024-04-01 paper.tex
removed latex H from table
small tweaks to notation and start on methodology

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$\ }
\newcommand\kD{$k\text{-}D$\ }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

At first glance, thinking of of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world''. The \pD observations are the realization of the
phenomenon, and the \kD NLDR layout is the simplified representation.
From a statistical perspective we can consider the distances between
points in the \kD layout to be variance that the model explains, and the
(relative) difference with their distances in \pD is the error, or
unexplained variance. Abstractly, we can also imagine that the
positioning of points in 2D represents fitted values, that will have
some prescribed position in \pD that can be compared with the observed
value.

\subsection{Notation}\label{notation}

Table~\ref{tbl-notation} summarises the notation used to explain the new
methodology.

@@

[H]
\centering\begingroup\fontsize{12}{14}\selectfont

\begin{tabular}{>{\raggedright\arraybackslash}p{5cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, d$ & number of observations, variables, embedding dimension, respectively\\
$X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$ & observed data matrix\\
$Y_{n \times d} = \begin{bmatrix} \textbf{y} _{1} & \textbf{y}_ {2} & \cdots & \textbf{y}_{n} \\  \end{bmatrix}^\top$ & reduced dimension, embedding matrix\\
$P_t$, $P_{t+1}$ & point along the geodesic path at time $t$ and the next time instant, which is $t+1$\\
$y_{max}$ & maximum value of the scaled second embedding component\\
\addlinespace
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
\addlinespace
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
\addlinespace
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$ & hexagonal bin centroid coordinates of the $k^{th}$ hexagon\\
$C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$ & $p$-D mappings of 2D hexagonal bin centroids of the $k^{th}$ hexagon\\
\bottomrule
\end{tabular}
\endgroup{}

@@

Once we have notation, we need to have some math that gives precision to
the paragraph above. (XXX Michael??)

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$\ }
\newcommand\kD{$k\text{-}D$\ }
\newcommand\dD{$d\text{-}D$\ }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

At first glance, thinking of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world''. The \pD observations are the realization of the
phenomenon, and the \kD NLDR layout is the simplified representation.
From a statistical perspective we can consider the distances between
points in the \kD layout to be variance that the model explains, and the
(relative) difference with their distances in \pD is the error, or
unexplained variance. Abstractly, we can also imagine that the
positioning of points in 2D represents fitted values, that will have
some prescribed position in \pD that can be compared with the observed
value.

Table~\ref{tbl-notation} summarises the notation used to explain the new
methodology.
\(X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top\)
represents the observed data matrix, that are \(n\) \pD vectors sampled
from some population.

@@

\centering\begingroup\fontsize{12}{14}\selectfont

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$X$ & $p$-dimensional data\\
$Y$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d-dimensional$ linear projection of $p$-dimensional data\\
$y_{max}$ & maximum value of the scaled second embedding component\\
\addlinespace
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
\addlinespace
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
\addlinespace
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

We need to have some math that gives precision to the paragraphs in the
previous section, defining mapping from \pD to \kD to get the
\kD layout. Might be written below, too (XXX Michael??)

To illustrate the method, we use simulated data. The S-curve is
generated as follows XXX Jayani

\subsection{\texorpdfstring{Generating the
\kD layout}{Generating the layout}}\label{generating-the-layout}

The usual procedures are followed to produce a \kD layout, using the
preferred method, t-SNE, UMAP, \ldots{} The data may need to be
pre-processed by standardising or transforming to principal components.
